Vizualizing Toronto Homess Shelter data Across a Global Pandemic

### 2022-02-06 outputs/paper/paper.Rmd
Final Commit

You can and should cross-reference sections and sub-sections. For instance, Section \@ref(data). R Markdown automatically makes the sections lower case and adds a dash to spaces to generate labels, for instance, Section \@ref(first-discussion-point).

abstract: "In addition to a global pandemic, homelessness and housing is a self apparent issue in the city of Toronto. Taking a stroll through certain areas in the downtown core, it's almost impossible to miss homeless tent cities and homeless people lined outside of shelters awaiting food or entry in the afternoon. Upon further examination of data collected from the Open Data Portal of the City of Toronto, it was easy to find interesting trends in homeless shelter data in the years 2020 and 2021. We found trends in the diminishing use of shelters by various demographic of homeless people and age groups. This paper attempts to visualize these trends as well as touch on some problems that one might experience in trying to extract real meaning from the data."

@@

In this Data Section \@ref(data), I will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data.

@@

## Data Processing
To process our data, we began by loading all the necessary libraries like opendatatoronto @citeODT whichwas used to load the data into R @citeR, a package called lubridate @citelubridate which will be used to process date data which later proves to be quite tricky. We also used load in libraries knitr

```{r, fig.pos = "!H",echo = FALSE}
no_all_population %>%
  ggplot(mapping = aes(x = month, y = ageunder16, color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count of Shelter Users Under Age 16",
          caption = "Use of shelter users under age 16 by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()
```

```{r bills, fig.pos = "!H", fig.cap="Bills of penguins", echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}

monthly_shelter_usage %>% 
  select(population_group, ageunder16, `age16-24`, `age25-44`, `age45-64`, age65over) %>% 
  slice(1:10) %>% 
  kable(
    caption = "First ten rows of a dataset of shelture usage",
    col.names = c("Population Group", "Aged less than 16 ", "Aged 16-24", "Aged 25-44", "Aged 45-64", "Aged 65"),
    digits = 1,
    booktabs = TRUE,
    linesep = "" 
    )
```

```{r , fig.cap = "There are no Indeginous data for 2020", echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}

shelter_usage_by_demographic <-
  monthly_shelter_usage %>%
  ggplot(mapping = aes(x = `date(mmm-yy)`, 
                       fill = population_group)) +
  geom_bar() +
  coord_flip() +
  theme_linedraw()

shelter_usage_by_demographic

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r}
ageunder16_overtime <- all_population %>%
  ggplot(aes(x = date,
             y = ageunder16)) +
  labs(x = "Date",
       y = "Under 16",
       caption = "Shelter use by those aged under 16 in 2020 and 2021") +
  geom_line() +
  theme_minimal()
ageunder16_overtime
```

```{r Usage of Age Groups By Year, fig.cap = " ", echo = FALSE, fig.width=10, fig.height=10}
library(patchwork)
ageunder16_overtime <- all_population %>%
  ggplot(aes(x = date,
             y = ageunder16)) +
  labs(x = "Date",
       y = "Under 16",
       caption = "Shelter use by those aged under 16 in 2020 and 2021") +
  geom_line() +
  theme_minimal()

@@

ageunder16_overtime /
  age_16to24_overtime/
  age_25to44_overtime /
  age_45to64_overtime /
  age_66_plus

Talk way more about it.

# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful.

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

# Discussion

# Limitations

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\appendix

# Appendix {-}

# Additional details

>>>>

@@

In this Data Section \@ref(data), I will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data. First of, (Table \@ref(fig:)) give us a glimpse of the data.

@@

## Data Processing
To process our data, we began by loading all the necessary libraries like opendatatoronto [@citeODT] which was used to load the data into R [@citeR], a package called lubridate [@citelubridate] which will be used to process date data which later proves to be quite tricky. We also used load in libraries knitr [@citeknitr], tidyverse [@citetidyverse], tidyr [@citetidyr], janitor [@citejanitor], and dplyr [@citetidyr] at some points in the process for data processing and pdf document generation.

After loading the data into our local environment and saving its as a cvs called raw.csv, we began to create subsets of the data with the desired variables. we created a new column called total which creates a count of the total number of users in the SMIS data base for each month, by each population group. This was done by simply summing up all the values for the number of people in each age group. We append this to the data set and then create a new data set called clean where we removed the variables which we did not desire like returned from housing and only kept the date, the age groups, the total and the population groups.

In addition to the totals column, I also needed a way to track the progression of data use over time. The raw data sent does provide a date column but this date deemed to be very problematic. The date in the data set is of class character and is in the form mmm-yy. For example, "Jan-20". Because the class is a charterer, I attempted to use the as.Date() function in the Lubridate package @citelubridate in order to convert the the date to class date, so that the visuals that we produce treat is as such and we can get the data in chronological order rather than in alphabetical order. R [@citeR] is very picky about how it likes its dates to be formatted and because there is no day data provided and we only have month and year, I needed to use two different work around to get chronological time data. The first was to add a Month column which is class numerical with values 1-24. These values represent the amount of months elapsed since the start of 2020. For example, March 2021 will have the month number 15. This way, we can organize the data chronologically.

Alternatively, we also added a day column where the day is the last day of every month (note 2020 was a leap-year and thus the last day in February 2020 is the 29th). We then created a column where the three values month, day and year where combined into a string with the proper formatting and then that date was converted using the as.Date( ) to a value of class date.

We also created 2 new data sets, no_all_population and all_population which removed the all population data as well as removed all non All Population data respectively. Each data set (raw.csv, clean.csv, all_pop.csv and no_all_pop.csv) are all available on GitHub through the link on the title page.

## Population, Frame or Sample
It is important to discuss the reach of our data. As previously mentioned, the City of Toronto funds services for people experience homelessness and the data we have acquired for this paper reflects all the data collected in the SMIS system in the years 2020 and 2021. This is important because it brings to light one of the first limitations of the data which we will discuss more in Section \@ref(limitations). Although the SMIS data is quite expansive with 323 data points, it fails to capture the entire population of homeless people in the city of Toronto as it does not take into consideration homeless people who sleep on the street or those only using services which are not government funded. Thus, in this paper, we are only deal with a sample of the population, but a big one nonetheless.

Now, lets have a look at some of the final data we will be working with. They are illustrated below in Table \@ref(tab:tableofdata)
```{r tableofdata, fig.cap = "Glimps of the some of the data we're working with",fig.pos = "!H",echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}
no_all_population %>% 
  select(population_group, ageunder16, `age16-24`, `age25-44`, `age45-64`, age65over, total, month) %>% 
  slice(1:10) %>% 
  kable(
    caption = "First ten rows of a dataset of shelture usage without All Population",
    col.names = c("Population Group", "Aged less than 16 ", "Aged 16-24", "Aged 25-44", "Aged 45-64", "Aged 65", "Total", "Month"),
    digits = 1,
    booktabs = TRUE,
    linesep = "" 
    )
```
In Table \@ref(tab:tableofdata), we notice that there are alot of zeros but nowhere in the data do we see *NA* values. The reason for the zeros is the definition of various demographics. As mentioned, a Youth is defined as only being able to be someone aged 16 to 24 so we expect not to be able to see any values for Youth under any columns other than that age range.

# Visualizing the Data and The Implications
## Visual 1
In this analysis, I am interested at looking at how the the CVODI-19 pandemic has affected the flow of homeless shelter usage over the years 2020 and 2021 for the various population and age groups. In this section, we will display and explain the various visuals we were able to meaningfully extract from the data and the implications we are able to extract from them.

Lets look at a graph which shows the usage of homeless shelters by people aged under the age of 16 over 2020 and 2021.
```{r graph1, fig.cap = "Shelter use by those aged under 16", echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}
ageunder16_overtime <- all_population %>%
  ggplot(aes(x = date,
             y = ageunder16)) +
  labs(x = "Date",
       y = "Under 16",
       caption = "Shelter use by those aged under 16 in 2020 and 2021") +
  geom_line() +
  theme_minimal()
ageunder16_overtime
```
In this graph (Figure \@ref(fig:graph1)), we can see that shortly after the start of 2020, the use of government funded shelters in the SMIS database have been rapidly declining until June 2021 where it started to increase just as fast as it declined. Although at a surface level, it might seem like something very positive that shelter usage is decreasing, the fact of the matter is that homeless people simply do not disappear at such a rapid rate as illustrated in Figure \@ref(fig:graph1). Perhaps this finding is unique to people under the age of 16. Lets take a look at similar graphs for all the other age groups and we will then talk about the implications of the findings. We will use the Patchwork package [@citepatchwork] to visualize all the data together.

```{r secondgraph, fig.cap = "All age groups over time.", echo = FALSE, fig.width=10, fig.height=10, message = FALSE, warning = FALSE}
library(patchwork)
ageunder16_overtime <- all_population %>%
  ggplot(aes(x = date,
             y = ageunder16)) +
  labs(x = "Date",
       y = "Under 16",
       caption = "Shelter use by those aged under 16 in 2020 and 2021") +
  geom_line() +
  theme_minimal()

@@

ageunder16_overtime /
  age_16to24_overtime/
  age_25to44_overtime /
  age_45to64_overtime /
  age_66_plus
```

Above in Figure \@ref(fig:secondgraph) we see all the data on the usage of SMIS homeless shelters across all age groups. Across all age groups, we notice a similar trend. After the start of the pandemic, there is a rapid decrease in shelter usage. Lets focus on this to begin.

## Implication 1

As I've previously noted, homeless people simply do not disappear at the rate which Figure \@ref(fig:secondgraph) suggests. Homeless shelters are known for often being crowded and not offering the best of living conditions. Although it is realistic sense that to offer the most amount of homeless people with shelter, capital restricts the quality of the service provided: there is an important tradoff between the quality and the quantity of the service provided. I believe that this element is exactly the cause of the rapid decline in shelter use at the start of the pandemic. Here is why. We have learned that the Corona virus and all of its variants are highly contagious and social distancing is one of the main ways that public health officials have advised the global population to try and reduce the spread of the virus. Because shelters can be quite crowded, especially in the winter months when the temperature outdoors drops, it would appear from the above graphs that even with the temperatures slightly increasing outdoors at the start of spring in march, many homeless people including those aged under 16 opted to sleep outdoors where although they might be defenseless against the cold, they are protecting themselves against a possible COVID-19 infection.

This is definitely tragic and one might wonder how can we remedy this situation. The answer can be found in the second half of the graph.

So far, what we have done is comment on the implications of the first half of the graph where shelter usage was at a rapid decline. Now, let's look closely at the second part of the graph.

In the second half of the graph, we notice the opposite trend taking place. The shelter usage increases seemingly just as fast as it declines. This is great to see but will makes us wonder what could have caused this transition at the midway point of 2021. In my estimation, vaccination and education has been the solution. Because a world class race began to take place, scientists and epidemiologists from around the globe were able to develop vaccines in very short period of time in contrast to historical standards. As such, vaccination was widely available for all those wanting to get vaccinated and from the visuals above, it would appear that a combination of cooling temperatures and peoples garnered trust in the effectiveness of vaccines lead to confidence being restored to homeless folk who are looking for somewhere to stay at night.

## Visuals 2

Moving on, lets see what implications we can draw when we separate the data by population group.

```{r noindigeous, fig.cap = "There are no Indeginous data for 2020", echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}

shelter_usage_by_demographic <-
  monthly_shelter_usage %>%
  ggplot(mapping = aes(x = `date(mmm-yy)`, 
                       fill = population_group)) +
  geom_bar() +
  coord_flip() +
  theme_linedraw()
shelter_usage_by_demographic
```
In Figure \@ref(fig:noindigeous), we notice that for all the months which are in 2020, there is one missing bar. The bar that is missing is the Indigenous population which is not a metric that was tracked in 2020 and only started being tracked in 2021.

## Implications 2
We have long know that most (if not all) of the land that we use in Canada to live and operate through our daily lives once belonged to the first nations that lived here. Although it is not as simple as this, the land belongs to them and they have often historically gotten the wrong side of the stick as it pertains to their treatment in society and the settlements that they have come to terms with, with the Federal government. As a result, many indigenous folk decide to leave their reserves, where there are high rates of murders, shootings and drug addiction, for a shot at better prospects in the cities. This is often not the case. Indigenous people of all ages find themselves in hard situation and in many cases, end up homeless. It is hard to understand how indigenous folk who have already suffered so much at the hands of the Canadian governments at all levels, have been omitted as a variable for all these years up until 2021.

## Visuals 3
```{r bypop,fig.cap = "There are no Indeginous data for 2020", echo = FALSE, message = FALSE, warning = FALSE, fig.height= 10, fig.width=7, fig.align = "center"}
library(patchwork)
ageunder16_overtime2 <- no_all_population %>%
  ggplot(mapping = aes(x = month, 
                       y = ageunder16, 
                       color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count",
          caption = "Count of shelter users aged 16 by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()

age_16to24_overtime2 <- no_all_population %>%
  ggplot(mapping = aes(x = month, 
                       y = `age16-24`, 
                       color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count",
          caption = "Use of shelter users under aged 16 to by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()

age_25to44_overtime2 <- no_all_population %>%
  ggplot(mapping = aes(x = month, 
                       y = `age25-44`, color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count",
          caption = "Count of shelter users aged 25 to by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()

age_45to64_overtime2 <- no_all_population %>%
  ggplot(mapping = aes(x = month, 
                       y = `age45-64`, color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count",
          caption = "Count of shelter users aged 45 to 64 by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()

age_66_plus2 <- no_all_population %>%
  ggplot(mapping = aes(x = month, 
                       y = age65over, color = population_group)) +
  geom_line() +
  labs(x = "Months from January 2020",
          y = "Count ",
          caption = "Count of shelter users under aged 65 and above by population group.") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  theme_minimal()

ageunder16_overtime2 /
  age_16to24_overtime2/
  age_25to44_overtime2 /
  age_45to64_overtime2 /
  age_66_plus2
```

## Implications 3

In Figure \@ref(fig:bypop), further implications can be drawn on the situation of the Indigenous homeless community. While in the other population groups, we see some change over time in the use of homeless shelters (Youth is the other exception as the value for youth will always be 0 except when the age group is 16 to 24 years old.). When we look at the Indigenous data, we find that in all age groups, there is little to no variation in the shelter usage. In my estimation, this is further reinforcing Visuals \@ref(fig:noindigeous) and the implications we drew from it. There has been a lot of systemic racism throughout Canadian institutions towards minorities like indigenous folk and the data shows this clearly.

# Limitations
In this limitations section, I want to focus on the data which is provided to us and have a  look into why it might not be easy to take the implications mentioned above at face value

## Uneven Bin width for Age Groups.
The first major flaw I've found in the data is that the bins that the age groups are recorded in are not all the same size. For example, the first bin has a size of 16 years, the second bin from ages 16 to 24 has width 9, the next two have width 19 with the last ranging in theory from 65 to infinity. Although it is possible to understand why the maintainers of the data choose to have a demarcation point of the bins at 65 and 16, it is very hard to understand why they opted to divide the interior portion into varying bin sizes.

We know in data science that when plotting bar charts, the bin width can have a big implication on the presentation of the data and the things that we can infer about its distribution. If the bin width is too small or too large, we might reduce or add characteristics to the data that aren't actually representative of it.

## What about the others?
The next and last limitation that we have when using this data as our only primary source is that we don't have any data on the homeless people who remain sleeping outdoors and on the homeless people who use shelter systems for assistance but not those funded by the government or upheld by the SMIS.

# Next Steps
Following on from the latter section, we can make some pretty good suggestions on ways to improve the data so that the findings we have will be less ambiguous and more ethical to use with confidence. First off, we can start by creating eve bins that homeless people checking into SMIS will have to answer. An even better alternative would be to collect the exact age and date of birth of homeless people (date of birth so that the age can be increased at the appropriate date every year). In this manner, the data scientist who gets a hold of the data will have the liberty to conduct his analysis and chose the bins which work best for the task he is working om, instead of being obstructed by barriers built into the data. This will not only ease the job of the data scientist but will also improve the reproducibility of his workflow and finding.

Next, the city might want to look at how it can improve and expand the SMIS system as to keep track of all the homeless shelters and services in the City of Toronto, as well as find ways to consider and count homeless folk who sleep on the street
